---
title: "Filter and collapse medulloblastoma OpenPBTA data"
output: 
  html_notebook:
    toc: true
    toc_float: true
date: 2020
---

_Adapted from [Childhood Cancer Data Lab, Alex's Lemonade Stand Foundation training materials](https://github.com/AlexsLemonade/training-modules/tree/ddc9c76e03aaf147fb2f011490ea6aec835de064)._
_See [this notebook](https://htmlpreview.github.io/?https://github.com/AlexsLemonade/training-modules/blob/ddc9c76e03aaf147fb2f011490ea6aec835de064/machine-learning/03-openpbta_PLIER.nb.html) in particular!_

## Set up

### Libraries

```{r}
library(tidyverse)
```

### Directories and files

#### Directories

```{r directories}
# Directory that holds the metadata and RNA-seq
data_dir <- file.path("..", "data")
```

#### Input files

```{r input_files}
# Metadata
histologies_file <- file.path(data_dir, 
                              "pbta-histologies-stranded-rnaseq.tsv.gz")
# VST transformed RNA-seq data
rnaseq_file <- file.path(data_dir, "pbta-vst-stranded.tsv.gz")
```

#### Output file

```{r output_files}
# Metadata file that only contains relevant columns and medulloblastoma 
# samples
medullo_histologies_file <- file.path(data_dir,
                                      "pbta-medullo-histologies.tsv.gz")
# RNA-seq file with duplicate gene identifiers collapsed
collapsed_data_file <- file.path(data_dir, "pbta-medullo-vst-collapsed.tsv.gz")
```

## Read in and prepare data

### Read in sample metadata and identify medulloblastoma samples

In the interest of reducing runtime and resource requirements during the workshop, we're going to subset the dataset to only the medulloblastoma samples.

```{r read_in_metadata}
# Read in metadata
histologies_df <- read_tsv(histologies_file)
```

Filter only to the medulloblastoma samples and columns we will use in the workshop.

```{r filter_to_medullo}
# Filter only to relevant samples and columns + write to file
medulloblastoma_df <- histologies_df %>%
  filter(short_histology == "Medulloblastoma") %>%
  select(Kids_First_Biospecimen_ID, 
         short_histology, 
         broad_histology,
         tumor_descriptor, 
         molecular_subtype) %>%
  write_tsv(medullo_histologies_file)

# Get biospecimen identifiers for medulloblastoma samples that we'll use to
# subset the RNA-seq data
medulloblastoma_samples <- medulloblastoma_df %>%
  pull(Kids_First_Biospecimen_ID)
```

### Read in expression data

Read in the VST transformed RNA-seq data.

```{r read_in_rnaseq}
rnaseq_df <- read_tsv(rnaseq_file)
```

The `gene_id` column concatenates two different gene identifiers: Ensembl gene IDs and gene symbols.
The Pathway-Level Information ExtractoR (PLIER) ([Mao *et al.* (2019)](https://doi.org/10.1038/s41592-019-0456-1)) method we'll use as part of the workshop requires gene symbols as rownames, so we will need to take some steps to prepare this data for use with PLIER.

First, we need to separate the gene symbols from the Ensembl gene identifiers.
Luckily, there's a tidyverse function that can separate one column into multiple columns called `tidyr::separate()`.

```{r separate_gene_ids}
symbol_rnaseq_df <- rnaseq_df %>%
  tidyr::separate(gene_id,  # Take the gene_id column
                  # Create two new columns called ensembl_id and gene_symbol
                  into = c("ensembl_id", "gene_symbol"),
                  # The values for these two columns are separated by _
                  sep = "_",
                  # Some gene symbols themselves contain _, so when that happens
                  # merge everything after the first _ into the gene_symbol
                  # column
                  extra = "merge")

# We no longer need the data frame we read in, so we'll remove it to save some
# memory
rm(rnaseq_df)
```

Alright, what does our new data frame look like?

```{r head_symbol_df}
head(symbol_rnaseq_df)
```

We won't need the column with the Ensembl identifiers anymore, so let's remove it.

```{r remove_ensg}
symbol_rnaseq_df <- symbol_rnaseq_df %>%
  select(-ensembl_id)
```

### Collapsing duplicate gene symbols

```{r any_duplicated}
any(duplicated(symbol_rnaseq_df$gene_symbol))
```

How many duplicated gene symbols are there?

```{r how_many_duplicated}
sum(duplicated(symbol_rnaseq_df$gene_symbol))
```

Relative to the total number of genes, there are not too many genes that are duplicated.

```{r add_row_means}
# mean transformed value across the entire cohort
symbol_rnaseq_df$mean_value <- rowMeans(symbol_rnaseq_df[, -1])
```

We'll use a few gene symbols that we've arbitrarily selected to follow along with these processing steps!

```{r duplicated_examples}
# Genes we will use to illustrate what happens during the collapsing duplicate
# process
duplicated_symbols <- c("DGCR5", "RABGEF1")

# Let's take a look at the mean values prior to any filtering
symbol_rnaseq_df %>% 
  filter(gene_symbol %in% duplicated_symbols) %>%
  select(gene_symbol, mean_value) %>%
  arrange(gene_symbol)
```

For a given duplicated gene symbol, we want to select the row with the highest average expression value.

```{r collapsed_rnaseq}
# There's sampling in the case of ties, so set a seed for reproducibility
set.seed(2020)

collapsed_rnaseq_df <- symbol_rnaseq_df %>%
  # For each set of rows that correspond to the same gene symbol
  group_by(gene_symbol) %>%
  # Select the single row with the highest value in the mean_value column
  top_n(1, mean_value) %>%
  # In the case of ties, where the mean values are the same, randomly pick one
  # row
  sample_n(1)
```

Check again to see if any duplicated gene symbols remain in the new collapsed data frame.

```{r any_duplicated_now}
any(duplicated(collapsed_rnaseq_df$gene_symbol))
```

Which row for our gene symbols of interest was retained in the collapsed data frame?

```{r which_retained}
collapsed_rnaseq_df %>% 
  filter(gene_symbol %in% duplicated_symbols) %>%
  select(gene_symbol, mean_value) %>%
  arrange(gene_symbol)
```

Now we're ready to remove the extraneous column with our mean values and only our medulloblastoma samples and write to file.

```{r write_collapsed}
collapsed_rnaseq_df %>%
  select(gene_symbol, tidyselect::all_of(medulloblastoma_samples)) %>%
  write_tsv(collapsed_data_file)
```

## Session Info

```{r}
sessionInfo()
```

