---
title: "Prepare PLIER model"
output: 
  html_notebook:
    toc: true
    toc_float: true
date: 2020
---

_Adapted from Childhood Cancer Data Lab, Alex's Lemonade Stand Foundation [training materials](https://github.com/AlexsLemonade/training-modules/tree/ddc9c76e03aaf147fb2f011490ea6aec835de064)._

## Set up

```{r}
# We'll need to set the seed for reproducibility
seed <- 2020
```

### Libraries

```{r}
library(magrittr)
library(readr)
library(PLIER)
```

### Directories and files

#### Directories

```{r}
# Directory that contains processed RNA-seq data
data_dir <- file.path("..", "data")

# We'll save the PLIER model to a directory called models
models_dir <- file.path("..", "models")
dir.create(models_dir, showWarnings = FALSE)
```

#### Input files

We've performed variance-stabilizing transformation on RNA-seq data, collapsed the duplicate gene identifiers, and filtered to only medulloblastoma samples.
See `README.md` and `01-filter-and-collapse.Rmd` for more information.

```{r}
rnaseq_file <- file.path(data_dir, "pbta-medullo-vst-collapsed.tsv.gz")
```

#### Output files

We're going to save the PLIER model to the `models` directory.

```{r}
plier_file <- file.path(models_dir, "pbta-medullo-plier.RDS")
```

## Read in RNA-seq data

```{r}
medulloblastoma_rnaseq_df <- read_tsv(rnaseq_file)

# We need to convert this to a matrix where the gene symbols are rownames
medulloblastoma_rnaseq_mat <- medulloblastoma_rnaseq_df %>%
  tibble::column_to_rownames("gene_symbol") %>%
  as.matrix()
```

## PLIER

### Pathway data

The `PLIER` package comes with different gene sets we can use as input during training.
The gene sets themselves are not really any different than other gene sets we might use for pathway analysis, but the data that comes in the package is in a format that is already compatible for use with PLIER.

```{r plier_data}
data("bloodCellMarkersIRISDMAP")
data("canonicalPathways")
data("svmMarkers")

# Combine the pathway data from PLIER
all_pathways <- PLIER::combinePaths(bloodCellMarkersIRISDMAP, 
                                    canonicalPathways,
                                    svmMarkers)
```

### Row normalize

`PLIER` comes with a built-in function to row normalize the expression data.
This is z-scoring the expression values for each gene.

```{r row_normalize, live = TRUE}
# Row normalize -- z-scoring the genes
medulloblastoma_zscore <- PLIER::rowNorm(medulloblastoma_rnaseq_mat)
```

### Run PLIER

```{r common_genes}
# Identify genes that are common to the pathway data and the z-score mat
common_genes <- PLIER::commonRows(all_pathways, medulloblastoma_zscore)
```

Now for the main PLIER function.

```{r run_plier}
plier_results <- PLIER::PLIER(data = medulloblastoma_zscore[common_genes, ],
                              priorMat = all_pathways[common_genes, ],
                              rseed = seed)
```

### Save to file

```{r}
plier_results %>% write_rds(plier_file)
```

## Session Info

```{r}
sessionInfo()
```

